### 2015-04-20 paper.tex
latex checkpoint

\documentclass{article}
\usepackage{url}
\usepackage{mathtools}
\usepackage{tikz}
\usetikzlibrary{arrows,positioning}
\begin{document}
\title{Ripple Consensus Review}
\author{Peter Todd}
\date{FIXME}
\maketitle

\section{Activities performed}

The Ripple consensus review investigation had four major activities associated
with it:

\begin{enumerate}

    \item Reviewed whitepapers and development documentation at ripple.com

    \item Reviewed existing third-party critism.

    \item Setup and run a full node.

    \item Reviewed C++ implementation, specifically the 0.27.4 tag.\footnote{git commit 92812fe7239ffa3ba91649b2ece1e892b866ec2a}

\end{enumerate}

\section{Overall architecture}

Ripple has diverged significantly from the original concept of a decentralized
network representing money explicitly as debt relationship between
parties.\ref{btcmag-introducing-ripple} While the concept of a debt
relationship still exists in the form of trust lines between participants, the
bulk of the codebase and developer documentation now focuses\footnote{For
    instance the ``Tutorials'' section of the Ripple website only explains how
    to create transactions that modify the ledger; there is almost no information
available on how to actually use the trustlines feature.} on the use and
maintenance of a global ledger of transactions and account balances.
Additionally a native currency, XRP, has been added, which is used for
anti-spam transaction fees\footnote{A small amount of XRP is irrovocably
destroyed for every transaction.}, as well as to serve as a universal currency.

\subsection{Hashing and serialization}

Though various parts of the Ripple user API's and networking use industry
standard serialization formats like JSON and Google Protobuf, for
consensus-critical functionality Ripple uses a custom tag-length-value
serialization and hashing scheme. This is not unexpected as industry standard
serialization schemes rarely, if ever, account for the need to create hash
digests from represented objects.

\begin{equation}
    \textit{Serialize}(\text{obj}) = t_0 n_0 d_0 + \hdots + t_n n_n d_n
\end{equation}

Hashing of objects generally uses the following scheme, implemented by
STObject::getHash(), resulting in a $256\text{bit}$ digest:

\begin{equation}
    H(\text{obj}) = \text{SHA512}(p + \text{Serialize}(\text{obj}))[0:256\text{bits}]
\end{equation}

The prefix $p$ is per-object-type, guaranteeing that objects of different types
will always have a different hash.\footnote{This is commonly known as
\emph{tagged hashing} in the literature.} For objects containing signatures,
such as transactions, there is a similar but separate \emph{signing hash}
implemented by the function STObject::getSigningHash(). Unlike the standard
hash, the signing hash does not serialize signature fields.

\subsection{Transactions}

Unlike Bitcoin, transactions increment and decrement account balances; they do
not directly consume the output of other transactions. To prevent replay
attacks accounts and transaction have sequence numbers; a transaction is only
valid if the Sequence number is exactly one greater than the last-valided
transaction from the same account. Additionally an optional AccountTxnID field
is available for use within transactions; the transaction is only valid if the
sending account's previously-sent transaction matches the provided hash.

Authentication is performed with a simple signature scheme; a scripting system
is not available, nor is multisignature support. To implement the latter a
highly complex single-purpose scheme\ref{ripple-wiki-multisign} has been
proposed.

\subsection{Ledger}

Information on the exact structure of the Ripple ledger is somewhat spotty.

The Ripple consensus algorithm starts with a Unique Node List.

\section{}

\bibliographystyle{plain}
\bibliography{paper}

\end{document}

### 2015-04-20 paper.tex
flesh out attacks

@@
 
+how exactly does the skiplist work?
+
+\subsection{Consensus}
+
 The Ripple consensus algorithm starts with a Unique Node List.
 
-\section{}
+what happens with UNL majority is not reached?
+
+\section{Attacks}
+
+reference c3 criteria?
+
+big picture with attacks is there is not a minimum cost to them unlike PoW
+
+evaluate attacks for denial-of-service and theft potential; evaluate minimum
+cost to carry out attack; possible minimum manpower to carry out attack
+
+Type of attack: DoS and/or theft
+
+Minimum cost: $0, $1k, $10k, $100k
+
+Scope: targetted, broad, global
+Duration: hours, days, weeks, indefinite - How long will it take to fix the bug?
+
+we don't attempt to put a dollar figure on harm done, as depends enormously on
+how people use ripple
+
+\subsection{Consensus Split}
+
+1) exploit consensus failure.
+
+DoS: Cost \$0, Scope: global, Duration: hours
+Theft: Cost \$1k, Scope: varies, Duration: hours
+
+not as synergistic with other attacks; obvious flaw that will be fixed;
+time-limited attack
+
+\subsection{Scalability}
+
+DoS: Cost \$100k, Scope: global, Duration: weeks
+
+simply produce a lot of txs, paid for via fees, either covertely or overtly.
+System will fail, guaranteed.
+
+\subsection{UNL: Jurisdictional}
+
+DoS/Theft: Cost \$100k, Scope: varies, Duration: indefinite
+
+centralization attack - legal attack against centralized UNL. note
+jurisdictional issues with delegating consensus to UNL
+
+biggest issue is how can ripple compete against systems without the global
+consensus req? if I'm a bank in russia, why use ripple when it opens me to UNL
+issues?
+
+\subsection{Software backdoor}
+
+DoS/Theft: Cost \$0, Scope: varies, Duration: varies
+
+mention unsigned code here
+
+\subsection{UNL: key theft}
+
+prep: Cost \$0, Scope: global, Duration: N/A
+
+Gain control of UNL signing keys through theft/compromise.
+
+\subsection{History simulation}
+
+Theft: Cost \$0, Scope: targetted, Duration: weeks
+
+from proof-of-stake terminology; simulated history is costless as no work
+expended.
+
+use backdoored software as example of how this can happen; also can happen with
+jurisdictional attacks
+
+if rippled sees two versions of history, what happens?
+
+rewrite history attack - note lack of cost to rewrite because no PoW; single
+implementation so likely we can hack all nodes at once
+
+proof-of-work has a hard bound on rewrite attacks, because can't break laws of
+physics
+
+can talk about how cost to do a simulation attack is PoW $/targets; talk about slasher rule
+
+\section{Synergistic Attacks}
+
+do a directed graph of likely synergistic attack sequences
 

### 2015-04-21 paper.tex
checkpoint

@@
 
+Claims from ripple: consensus protocol will maintain correctness so long as $f
+\le (n-1)/5$ where $f$ is the number of Byzantian failures. However default UNL
+list is just $5$ entries, thus $f \le (5-1)/5 = 0.8$ - no failures can be
+tolerated.
+
+Put antoher way, you only need to compromise $20\%$ of the nodes in the UNL to
+censor a transactions/freeze funds. "failure to agree is agreement to defer"
+
+Is there incentive to use anything but the default UNL? Remember that if you
+choose a UNL more than $80\%$ different than others, you can be attacked on
+that basis; your full node determines validity of transactions anyway. e.g. the
+"add 1000 validators to UNL list" suggestion.
+
+mixing structure: if tx A is rejected, dependent txs are all rejected; will
+likely soon result in situation where you can't use different UNL
+
 The Ripple consensus algorithm starts with a Unique Node List.
@@
 what happens with UNL majority is not reached?
+
+test: add extra entries to UNL list, see if ledgers close, if they don't think we know threshold
 

### 2015-04-24 paper.tex
consensus discussion and diagrams

@@
 network representing money explicitly as debt relationship between
-parties.\ref{btcmag-introducing-ripple} While the concept of a debt
+parties.\cite{btcmag-introducing-ripple} While the concept of a debt
 relationship still exists in the form of trust lines between participants, the
@@
 is not available, nor is multisignature support. To implement the latter a
-highly complex single-purpose scheme\ref{ripple-wiki-multisign} has been
+highly complex single-purpose scheme\cite{ripple-wiki-multisign} has been
 proposed.
@@
 
-\subsection{Consensus}
+\subsection{Unique Node List}
 
-Claims from ripple: consensus protocol will maintain correctness so long as $f
+The actual process of determining what version of the ledger is correct starts
+with the Unique Node Listi, (UNL) a list of public keys meant to be associated
+with active nodes the node operator believes are
+``unique''.\cite{ripple-wiki-unl} Ripple Labs suggests that UNL's ``should have
+100+ nodes on them.''\cite{ripple-wiki-unl} and provides a ``starter'' UNL at
+\url{https://ripple.com/ripple.txt} with $5$ to $8$ nodes.\footnote{Exactly how this UNL is
+generated is unknown; the author downloaded it on multiple occasions getting
+between $5$ to $8$ nodes each time, mostly the same.}
+
+Through the Ripple Protocol Consensus Algorithm\cite{ripple-consensus-paper}
+nodes on the UNL vote to determine the contents of the Ripple ledger. While the
+actual protocol contains a number of rounds of proposals and voting the end
+result can be described as basically a supermajority vote: a transaction is
+only approved if $80\%$ of the UNL of a server agrees with
+it.\cite[3.2]{ripple-consensus-paper} Put another way, if $20\%$ of the UNL
+choose to reject a transaction, it will not be included in the ledger.
+
+Different nodes may chose to use different UNLs; if the UNLs do not
+sufficiently overlap global consensus is not guaranteed because different UNL
+``cliques'' can come to consensus independently of each other. For instance
+Figure \ref{fig:unl-clique-disjoint} shows a simple example with two UNL
+cliques, red and blue, that have forked because of insufficient connectivity,
+resulting in the forked blockchain shown in Figure \ref{fig:unl-clique-disjoint-blockchain}.
+
+Increasing the connectivity past the $20\%$ threshold, as shown in Figure
+\ref{fig:unl-clique-consensus}, brings both cliques into agreement. However
+what happens to the blockchain? The Ripple Protocol Consensus Algorithm paper
+is unclear on this point. For instance, the loss and restoration of
+connectivity may have been associated with network latency. The paper does
+state that "nodes who’s latency grows larger than a preset bound $b$ are
+removed from all UNLs"\cite[3.4.1]{ripple-consensus-paper} but does not say if
+this removal is meant to be permanent\footnote{In the existing implementation
+restarting the node does reload the UNL set from the config file.} nor if by
+``all'' they refer to removal from UNLs on all nodes in the network.
+Additionally while the author did not have sufficient time to investigate this
+point fully, the source code doesn't appear to have any provision for
+reorganizations of the ledger when a longer fork is detected.
+
+Regardless, it is certain that following a fork caused by disjoint UNL cliques
+the losing side simply has to discard some or all history, as seen in Figure
+\ref{fig:unl-clique-post-fork}, possibly resulting in double-spends and
+associated financial losses.
+
+\begin{figure}
+    \centering
+    \includegraphics{figures/unl-clique-disjoint.eps}
+    \caption{Disjoint UNL}
+    \label{fig:unl-clique-disjoint}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics{figures/unl-clique-disjoint-blockchain.eps}
+    \caption{Forked blockchain}
+    \label{fig:unl-clique-disjoint-blockchain}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics{figures/unl-clique-consensus.eps}
+    \caption{UNL cliques in consensus}
+    \label{fig:unl-clique-consensus}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics{figures/unl-clique-reorganized-blockchain.eps}
+    \caption{Post-fork reorganization}
+    \label{fig:unl-clique-post-fork}
+\end{figure}
+
+Similarly for a UNL of $n$ nodes it would
+take $(4n + 1)/5$ Byzantine failures for an incorrect transaction to be
+confirmed.
+
+Claims from ripple: consensus protocol\cite{ripple-consensus-paper} will maintain correctness so long as $f
 \le (n-1)/5$ where $f$ is the number of Byzantian failures. However default UNL
@@
 
-can talk about how cost to do a simulation attack is PoW $/targets; talk about slasher rule
+can talk about how cost to do a simulation attack is PoW targets; talk about slasher rule
 

### 2015-04-25 paper.tex
Choosing the Unique Node List

@@
 \usepackage{tikz}
+\usepackage{multirow}
 \usetikzlibrary{arrows,positioning}
@@
 cliques, red and blue, that have forked because of insufficient connectivity,
-resulting in the forked blockchain shown in Figure \ref{fig:unl-clique-disjoint-blockchain}.
+resulting in the forked blockchain shown in Figure
+\ref{fig:unl-clique-disjoint-blockchain}. More complex failure modes have also
+been identified, such as the ``less than majority evil'' failure mode
+identified by Gregory Maxwell,\cite{gmaxwell-btctalk-ripple} and ``separate
+supermajorities'' mode identified by Andrew Miller.\cite{amiller-rippletalk}
 
@@
 Regardless, it is certain that following a fork caused by disjoint UNL cliques
-the losing side simply has to discard some or all history, as seen in Figure
-\ref{fig:unl-clique-post-fork}, possibly resulting in double-spends and
+the losing side simply has to discard some or all history (Figure
+\ref{fig:unl-clique-post-fork}) possibly resulting in double-spends and
 associated financial losses.
@@
 
-Similarly for a UNL of $n$ nodes it would
-take $(4n + 1)/5$ Byzantine failures for an incorrect transaction to be
-confirmed.
+\subsection{Choosing the Unique Node List}
+
+How should Ripple node administrators pick their Unique Node List? Ripple Labs
+provides little, if any, concrete advice beyond the ``starter UNL'' they
+provide. Let's look at this problem from another angle: Why shouldn't a node
+operator just use the default UNL?
+
+Let's assume the node operator does not know what UNL other nodes in the
+economic majority have chosen to use. They do however know the default UNL
+provided by Ripple Labs, and they have a reasonable belief that other node
+operators also know that UNL and may also be using it. The node operator's goal
+is to limit losses; the best case is their node stays in consensus with other
+nodes. Second best is they suffer a denial of service attack causing their node
+to stop processing some or all transactions. Worst case is their node accepts
+transactions that other nodes do not, making possible double-spend attacks.
+Figure \ref{fig:unl-choice-plot} shows the outcomes of various possible
+decisions by them and other node operators, ranging from $100\%$ default UNL,
+$50\%$ default UNL, and $10\%$ default UNL.
+
+In every circumstance node operators can reduce the risk of losses due to
+consensus failures by removing non-default UNL entries, with the least risky
+option being to simply stick with the default UNL of $100\%$ Ripple Labs
+controlled nodes. This is particularly acute given the low profit margins of
+many financial transactions: while a denial-of-service halts incoming revenue,
+a double-spend attack can quickly drain working capital, quite possibly without
+any hope of ever recovering the stolen funds.
+
+A closely related issue is what incentive does a node operator have to publicly
+perform validation services? Ripple currently has no compensation mechanism for
+public validators, yet validation at minimum raises potential legal issues such
+as lawsuits for negligence. Again, the least-risk option is to not publicly
+validate the Ripple ledger.
+
+\begin{table}
+    \begin{tabular}{c|c|c|c|}
+        ~ & \multicolumn{3}{c}{Majority UNL \%} \\
+        \multirow{3}{*}{You} & $100\%$ & $50\%$ & $10\%$ \\ \hline
+        $100\%$ & Consensus & DoS & DoS \\ \hline
+        $50\%$ & DoS & DoS & D.S. \\ \hline
+        $10\%$ & D.S & D.S & D.S. \\
+    \end{tabular}
+\end{table}
 

### 2015-04-26 paper.tex
wrote up some attacks

@@
 \usepackage{tikz}
-\usepackage{multirow}
+% \usepackage{multirow}
 \usetikzlibrary{arrows,positioning}
@@
         ~ & \multicolumn{3}{c}{Majority UNL \%} \\
-        \multirow{3}{*}{You} & $100\%$ & $50\%$ & $10\%$ \\ \hline
+        % \multirow{3}{*}{You} & $100\%$ & $50\%$ & $10\%$ \\ \hline
         $100\%$ & Consensus & DoS & DoS \\ \hline
-        $50\%$ & DoS & DoS & D.S. \\ \hline
-        $10\%$ & D.S & D.S & D.S. \\
+        $50\%$ & DoS & DoS & Doublespend \\ \hline
+        $10\%$ & Doublespend & Doublespend & Doublespend \\
     \end{tabular}
@@
 
-Claims from ripple: consensus protocol\cite{ripple-consensus-paper} will maintain correctness so long as $f
-\le (n-1)/5$ where $f$ is the number of Byzantian failures. However default UNL
-list is just $5$ entries, thus $f \le (5-1)/5 = 0.8$ - no failures can be
-tolerated.
+\section{Attack scenarios}
 
-Put antoher way, you only need to compromise $20\%$ of the nodes in the UNL to
-censor a transactions/freeze funds. "failure to agree is agreement to defer"
+To evaluate potential attacks on the Ripple network we look at three factors:
 
-Is there incentive to use anything but the default UNL? Remember that if you
-choose a UNL more than $80\%$ different than others, you can be attacked on
-that basis; your full node determines validity of transactions anyway. e.g. the
-"add 1000 validators to UNL list" suggestion.
+\begin{itemize}
 
-mixing structure: if tx A is rejected, dependent txs are all rejected; will
-likely soon result in situation where you can't use different UNL
+    \item \textbf{Type:} What is the effect of the attack? Denial of Service
+          and/or theft? Or is the attack mostly useful in preparation for another
+          type of attack?
 
-The Ripple consensus algorithm starts with a Unique Node List.
+    \item \textbf{Cost:} What is a probable minimum cost to carry out the attack, for
+          the people able to do the attack? For instance, a technical attack
+          that can be launched by ``bored teenagers'' will have an essentially
+          zero cost while a long-term rewrite of a proof-of-work blockchain has
+          a reasonably well-defined cost.
 
-what happens with UNL majority is not reached?
+    \item \textbf{Scope:} Who is affected by the attack? Is this a targetted attack,
+          mainly affecting a one or two targets? An attack with broad impact? Or
+          an attack with global impact on all Ripple users?
 
-test: add extra entries to UNL list, see if ledgers close, if they don't think we know threshold
+    \item \textbf{Duration:} How long will it take before the attack is
+          neutralized?  This may be a matter of hours to write and distribute a
+          simple software patch\footnote{The author personally stopped the
+          CVE-2013-4627 attack on the Bitcoin network in a few hours of rushed
+          work in the middle of the night.}, to indefinitely long for some
+          attacks where recovery is impossible.
 
-\section{Attacks}
+    \item \textbf{Probability:} How likely is it for the Ripple network to be
+          attacked this way?
 
-reference c3 criteria?
+\end{itemize}
+
+We do \emph{not} try to determine the monetary impact of the attack, because we
+are unable to predict how the Ripple network will be used. For instance, while
+a double-spend attack may have the same basic impact on a $\$100/\text{day}$
+small business as it would on a $\$1,000,000,000/\text{day}$ multinational, the
+latter could lose orders of magnitude more money for what it at a technical
+level basically the same attack. A closely issue with attempting to predict
+monetary losses is that at some point losses can be sufficiently high that they
+create situations where social consensus can decide that the losses simply must
+be reversed. For instance, Vericoin community chose to do a hard-fork of their
+alt-currency to undo the impact of a major theft.\cite{coindesk-vericoin}
 
@@
 
-1) exploit consensus failure.
+The attacker exploits a difference in behavior between different
+implementations/versions of the Ripple protocol. The result of the attack can
+be a simple denial of service due to the Ripple network being unable to process
+transactions, or with more sophistication, the attacker can fork the Ripple
+ledger/blockchain and use the fork to doublespend. The attack is stopped by
+identifying it, and then distributing software patches. Provided the Ripple
+network is well monitored consensus split attacks can be stopped fairly
+quickly; the March 2013 Bitcoin chain fork\cite{bip50} was resolved in a matter
+of hours by contacting mining pools. A similar response could happen on Ripple
+by contacting validating nodes commonly used in UNLs.\footnote{Another reason
+to use the UNL provided Ripple Labs.}
 
-DoS: Cost \$0, Scope: global, Duration: hours
-Theft: Cost \$1k, Scope: varies, Duration: hours
+Note that a consensus split may also happen by accident; prior consensus splits
+in cryptocurrency systems have almost always been accidental. Equally, that
+consensus splits can happen by accident clearly shows that the minimum cost to
+perform this attack is zero. However, due to the limited number of consensus
+bugs available to exploit in a given (set of) versions of the Ripple protocol
+in use and the fast response time possible to consensus splits the potential of
+this class of attack for long-term attacks on the network is limited.
+
+Risk factors:
+
+\begin{itemize}
+
+    \item The main implementation of the Ripple protocol, \emph{rippled}, does
+        not cleanly separate the consensus-critical part of the codebase from
+        the non-consensus-critical part.\footnote{The consensus-critical
+            portions of Bitcoin Core codebase are currently being separated
+            into a stand-alone \emph{libconsensus} library.}
+
+    \item The Ripple protocol itself is extremely complex, with many different
+          types of transactions, and all functionality being implemented directly
+          in the protocol itself.
+
+    \item Improvements to Ripple frequently require changing the
+          consensus-critical codebase rather than user software. For instance
+          while on Bitcoin the implementation of multisig was possible without
+          modification to the protocol\cite{bip19} in Ripple the lack of
+          extension capabilities such as scripting require a consensus-critical
+          change.\cite{ripple-wiki-multisign}\footnote{P2SH was \emph{not}
+          required for multisig to be implemented.}
+
+\end{itemize}
+
+DoS: Cost \$0, Scope: global, Duration: hours, Prob: high
+Theft: Cost \$1k, Scope: varies, Duration: hours, Prob: high
 
@@
 
-\subsection{Scalability}
+\subsection{Transaction Flood}
 
-DoS: Cost \$100k, Scope: global, Duration: weeks
+DoS: Cost depends, Scope: broad, Duration: depends
 
-simply produce a lot of txs, paid for via fees, either covertely or overtly.
-System will fail, guaranteed.
+The attacker creates large numbers of transactions - either covertely or
+overtly - until the Ripple network fails due to overload, or more likely,
+Ripple users are priced out of their ability to use the Ripple network.
+Transactions pay a minimum fee per transaction by destroying XRP; the fee per
+transaction is set by a vote between all validators on the UNL list. In the
+short to medium term such an attack would quickly drive up that fee. A
+succesful attack requires deep pockets as the attacker must outspend other
+Ripple users.
+
+Risk factors:
+
+\begin{itemize}
+
+    \item Ripple Labs plans\cite{cointelegraph-ripple-aml} to require verified
+        user identification to use the Ripple network in the near future.  When
+        implemented this will make it easy to determine who is doing a covert
+        transaction flood attack and restrict their access to the network.
+
+    \item The global consensus ledger has inherently poor $O(n^2)$ scaling;
+        even in the absense of a malicious attack the poor scalability of the
+        ledger is a threat to the viability of Ripple.
+
+    \item A clever attacker wishing to overtly attack the Ripple network can
+        masquerade their traffic as legit economic activity; from the point of
+        view of Ripple Labs - a major XRP owner - such an attack may not even
+        be considered an attack!\footnote{Similar to the debates in the Bitcoin
+        ecosystem about whether or not fee-paying uses of the Bitcoin
+        blockchain for non-Bitcoin-denominated transactions constitute
+        ``attacks''.}
+
+    \item Attacking the poor scalability of Ripple is highly synergistic with
+        attacks depending on the high centralization of the Ripple network as
+        even ``unsuccesful'' scalability attacks drive further centralization.
+        For instance, if the response to a scalability attack is for Ripple
+        Labs to beef up the validation nodes they control, and users stop
+        running full nodes themselves, the network is more vulnerable to any
+        attacks on those centrally managed validation nodes.
+
+\end{itemize}
 
@@
 Gain control of UNL signing keys through theft/compromise.
+
+\subsection{Validator Fear, Uncertainty, and Doubt}
+
+make people afraid to run validating nodes
+
+illegal data
 

### 2015-04-26 paper.tex
Fix UNL outcomes table formatting

@@
 \usepackage{url}
+% \usepackage{showframe}
 \usepackage{mathtools}
 \usepackage{tikz}
-% \usepackage{multirow}
+\usepackage{multirow}
 \usetikzlibrary{arrows,positioning}
@@
 transactions that other nodes do not, making possible double-spend attacks.
-Figure \ref{fig:unl-choice-plot} shows the outcomes of various possible
+Table \ref{tbl:unl-outcomes} shows the outcomes of various possible
 decisions by them and other node operators, ranging from $100\%$ default UNL,
@@
 \begin{table}
-    \begin{tabular}{c|c|c|c|}
-        ~ & \multicolumn{3}{c}{Majority UNL \%} \\
-        % \multirow{3}{*}{You} & $100\%$ & $50\%$ & $10\%$ \\ \hline
-        $100\%$ & Consensus & DoS & DoS \\ \hline
-        $50\%$ & DoS & DoS & Doublespend \\ \hline
-        $10\%$ & Doublespend & Doublespend & Doublespend \\
+    \centering
+
+    \begin{tabular}{cc|c|c|c}
+        \multicolumn{2}{c}{} & \multicolumn{3}{c}{Majority} \\
+        \multicolumn{2}{c}{} & $100\%$ & $50\%$ & $10\%$ \\ \cline{3-5}
+        \multirow{3}{*}{Local} & $100\%$ & Consensus & DoS & DoS \\ \cline{2-5}
+        ~ & $50\%$ & DoS & DoS & Doublespend \\ \cline{2-5}
+        ~ & $10\%$ & Doublespend & Doublespend & Doublespend \\
     \end{tabular}
+
+    \caption{Worst-case outcomes of default UNL \% choices}
+    \label{tbl:unl-outcomes}
 \end{table}

### 2015-04-26 paper.tex
More fleshed out attacks

@@
 \subsection{Choosing the Unique Node List}
+\label{choosing-the-unl}
 
@@
 
-    \item \textbf{Scope:} Who is affected by the attack? Is this a targetted attack,
+    \item \textbf{Scope:} Who is affected by the attack? Is this a targeted attack,
           mainly affecting a one or two targets? An attack with broad impact? Or
@@
 
-big picture with attacks is there is not a minimum cost to them unlike PoW
+\subsection{Consensus Split}
 
-evaluate attacks for denial-of-service and theft potential; evaluate minimum
-cost to carry out attack; possible minimum manpower to carry out attack
-
-Type of attack: DoS and/or theft
-
-Minimum cost: $0, $1k, $10k, $100k
-
-Scope: targetted, broad, global
-Duration: hours, days, weeks, indefinite - How long will it take to fix the bug?
-
-we don't attempt to put a dollar figure on harm done, as depends enormously on
-how people use ripple
-
-\subsection{Consensus Split}
+\begin{center}
+    \begin{tabular}{c|c|c|c|c}
+        Type & Cost & Scope & Duration & Probability \\ \hline
+        DoS   & $\$0$ & Global & Hours & Very High \\
+        Theft & $\$1\text{k}$ & Targeted & Hours & Medium
+    \end{tabular}
+\end{center}
 
@@
 
-DoS: Cost \$0, Scope: global, Duration: hours, Prob: high
-Theft: Cost \$1k, Scope: varies, Duration: hours, Prob: high
-
-not as synergistic with other attacks; obvious flaw that will be fixed;
-time-limited attack
-
 \subsection{Transaction Flood}
 
-DoS: Cost depends, Scope: broad, Duration: depends
+\begin{center}
+    \begin{tabular}{c|c|c|c|c}
+        Type & Cost & Scope & Duration & Probability \\ \hline
+        DoS & $>\$100\text{k}$ & Broad & Varies & Low
+    \end{tabular}
+\end{center}
 
@@
 
-\subsection{UNL: Jurisdictional}
+\subsection{Coercion of validators}
 
-DoS/Theft: Cost \$100k, Scope: varies, Duration: indefinite
+\begin{center}
+    \begin{tabular}{c|c|c|c|c}
+        Type & Cost & Scope & Duration & Probability \\ \hline
+        DoS & $>\$100\text{k}$ & Targeted & Indefinite & High \\
+        DoS & $>\$1\text{M}$ & Global & Indefinite & Medium \\
+        Theft & $>\$100\text{k}$ & Targeted & Indefinite & Medium \\
+        Theft & $>\$1\text{M}$ & Global & Indefinite & Low
+    \end{tabular}
+\end{center}
 
-centralization attack - legal attack against centralized UNL. note
-jurisdictional issues with delegating consensus to UNL
+Validators are coerced through poltical/legal/criminal means into attacking
+some or all Ripple users by preventing valid transactions from entering the
+ledger (DoS) and/or by allowing invalid transactions to enter the ledger.
+(theft) The cost estimates above are very rough, and assuming the attack is via
+legal means; the cost may potentially be as low as zero in some scenarios like
+blackmail. In preparation for this attack the attackers may also attempt to
+reduce the number of validators by with, for instance, a transaction flood
+attack.
 
-biggest issue is how can ripple compete against systems without the global
-consensus req? if I'm a bank in russia, why use ripple when it opens me to UNL
-issues?
+Risk factors:
+
+\begin{itemize}
+
+    \item The default UNL list and incentives against operating or using
+        non-official validators discussed in Section \ref{choosing-the-unl}
+        significantly increase the effectiveness of a coercion attack as nearly
+        all Ripple users will be affected. Equally, in the event of a coercion
+        attack there are still significant downsides to adopting a different
+        UNL.
+
+    \item Similarly this attack is synergistic with transaction flooding and
+        other attacks on the scalability of the network, reducing the number of
+        validators that need to be coerced.
+
+    \item The rippled node implementation does not check full history. This
+        means if validators sign an invalid ledger, the transactions that make
+        it invalid can be hidden from Ripple users. Equally, even with
+        detection, the fact that Ripple users can be asked to simply
+        restart/reinstall their nodes to get past the invalid history may
+        significantly reduce the social barriers to creating an invalid ledger.
+
+    \item The Ripple protocol is constructed such that it is possible to create
+        many, although not all, types of fraud
+        proofs\cite{irc-btcdev-gmaxwell-fraud-proof, btctalk-pkt-fraud-proofs}
+        required to prove an invalid ledger. While the software does not
+        currently implement fraud proofs, this is functionality that could be
+        added in the future without a change to the consensus. If widely used
+        fraud proofs could prevent some types of coercion attacks that force
+        validators to create invalid ledgers by making it likely for the result
+        of the attack to be DoS rather than profitable theft. In short, the
+        attack would simply shut the Ripple network down for all users rather
+        than result in a succesful theft.
+
+    \item Some people would argue that the recently announced
+        plans\cite{cointelegraph-ripple-aml} to require verified user
+        identification to use the Ripple network in the near future are an
+        example of a succesful coercion attack. It can be argued that Ripple
+        Labs is increasing the risk of coercion attacks by simultaneously
+        adopting a ``regulator friendly'' position and promoting the use of
+        Ripple across many different jurisdictions.
+
+\end{itemize}
 
@@
 
-DoS/Theft: Cost \$0, Scope: varies, Duration: varies
+\begin{center}
+    \begin{tabular}{c|c|c|c|c}
+        Type & Cost & Scope & Duration & Probability \\ \hline
+        Varies & $\$0$ & Broad & Weeks & High \\
+    \end{tabular}
+\end{center}
 
-mention unsigned code here
+The attacker inserts a backdoor into a widely used Ripple protocol
+implementation. (e.g. the rippled node software) The attacker may be an
+internal Ripple Labs employee, an external member of the public submitting a
+pull-request, or they may compromise the software actually downloaded by
+end-users by compromising binary or sourcecode hosting. (e.g. GitHub)
 
-\subsection{UNL: key theft}
+Risk factors:
 
-prep: Cost \$0, Scope: global, Duration: N/A
+\begin{itemize}
 
-Gain control of UNL signing keys through theft/compromise.
+    \item All relevant software is open source, allowing for third-party
+        review. Ripple Labs also has a bug bounty program, which can
+        incentivise that review.
 
-\subsection{Validator Fear, Uncertainty, and Doubt}
+    \item Ripple Labs appears to use GitHub - a third party -
+        as their master sourcecode repository, with development happening
+        through pull-requests on GitHub. While this does place trust in a
+        third-party, PGP signing sourcecode during development can mitigate
+        this risk. However there appears to be no PGP signatures on any git
+        commits, giving Ripple Labs little ability to audit who wrote what
+        code.
 
-make people afraid to run validating nodes
+    \item \textbf{Currently Ripple Labs does not provide a secure way to
+        download any of their software.} While binaries are available, the
+        mechanism to download them is a Ubuntu package repository over insecure
+        HTTP; while source code is available through git, neither commits nor
+        release tags are signed in anyway. \textbf{This is a serious ommission
+        that has lead to significant monetary losses in the past.} Ripple Labs
+        should be following industry best-practice by signing git commits and
+        tags\cite{wladimir-git-pgp} as well as PGP signing their Ubuntu
+        packages.
 
-illegal data
+\end{itemize}
 
-\subsection{History simulation}
+\subsection{Theft of validator secret keys}
 
-Theft: Cost \$0, Scope: targetted, Duration: weeks
+\begin{center}
+    \begin{tabular}{c|c|c|c|c}
+        Type & Cost & Scope & Duration & Probability \\ \hline
+        Precursor & $\$0$ & N/A & N/A & Unknown \\
+    \end{tabular}
+\end{center}
 
-from proof-of-stake terminology; simulated history is costless as no work
-expended.
+The attacker steals illegitimate control of the secret keys used by validators,
+particularly the validators in the default UNL. The theft may happen via an
+exploit, backdoor, compromised hosting company, etc. While not useful for the
+attacker in of itself, the stolen keys can be used for further atttacks such as
+DoS or a simulated ledger attack.
 
-use backdoored software as example of how this can happen; also can happen with
-jurisdictional attacks
+A closely related attack is to compromise the Ripple website, and trick people
+into downloading a default UNL list consisting of only attacker controlled
+nodes.
 
-if rippled sees two versions of history, what happens?
+Risk factors:
 
-rewrite history attack - note lack of cost to rewrite because no PoW; single
-implementation so likely we can hack all nodes at once
+\begin{itemize}
 
-proof-of-work has a hard bound on rewrite attacks, because can't break laws of
-physics
+    \item Due to the high centralization of UNL validators, a relatively small
+        number of validators need to be compromised for a succesful attack.
 
-can talk about how cost to do a simulation attack is PoW targets; talk about slasher rule
+    \item Most validators are running nearly the same, if not the exact same,
+        software, making it likely for backdoor/exploit attack to succesfully
+        exfiltrate the secret keys. Due to the high consensus-critical
+        complexity of the Ripple protocol and lack of a consensus library
+        alternate validator implementations are currently infeasible, making it
+        difficult to change this situation.
+
+    \item The rippled implementation doesn't yet appear to support separating
+        the validation/consensus and signing functionality into different
+        machines, or even different user accounts. (possibly Ripple Labs has
+        internal-use-only improvements)
+
+    \item Details about the operational security practices of the validators on
+        the default UNL is unavailable, making it difficult to assess risk.
+
+    \item The Ripple P2P protocol does make it possible to operate validators
+        without the nodes themselves accepting public traffic; the signatures
+        produced by the validating nodes are distributed via an anonymizing
+        flood-fill network. This setup is recommended by Ripple Labs
+        documentation and makes remote attacks significantly more difficult.
+
+\end{itemize}
+
+\subsection{Simulated ledger}
+
+\begin{center}
+    \begin{tabular}{c|c|c|c|c}
+        Type & Cost & Scope & Duration & Probability \\ \hline
+        Theft & $>\$1\text{k}$ & Targeted & Days & N/A
+    \end{tabular}
+\end{center}
+
+An attacker with the ability to create signatures from nodes on the UNL can
+create an entirely simulated ledger in parallel to the ``official'' Ripple
+ledger, containing any transaction the attacker wishes. In conjunction with
+sybil attacking the target the attacker can then perform double-spend attacks.
+Sybil attacks are relatively cheap, requiring just the aquisition of a large
+number of IP addresses, e.g. via cloud hosting and/or botnet rental. In some
+situations sybil attacks can even have zero cost, such as the internal IT staff
+within the target with access to network infrastructure, the target's ISP, or
+government actors with MITM capabilities such as the
+NSA.\cite{schneier-nsa-mitm}
+
+Risk factors:
+
+\begin{itemize}
+
+    \item Because the Ripple consensus is determined by signatures creating a
+        simulated ledger is a zero-cost attack, the ``costless
+        simulation''\cite[4.2]{apoelstra-on-stake} problem. Proof-of-work
+        ledgers in comparison have very well-defined costs that discourage
+        attacks and incentivise detection.\footnote{On the Bitcoin network even
+            if the attacker can steal the hashing power needed to simulate
+        history for free the miners involved lose money from the attack and
+    have strong monetary incentives to detect and stop it.}
+
+    \item There are no direct incentives built into the Ripple protocol to
+        discourage validators from allowing their private keys to sign
+        simulated ledgers. By comparison the Slasher proof-of-stake algorithm
+        attempts to disincentivise simulated history by destroying valuable
+        bonds owned up by participants in the consensus
+        process.\cite{ethereum-blog-slasher}
+
+    \item The lack of incentive may invite coercion attacks, as the validators
+        being coerced into signing a simulated ledger have no direct incentive
+        not too. For instance a court could order a validator to assist in a
+        simulation attack on a target as a way of recovering funds held by the
+        target.
+
+    \item Simulation attacks are made significantly easier by the fact that the
+        Ripple consensus protocol will remove validators from the UNL if they
+        appear to have failed.\cite[3.4.2]{ripple-consensus-paper} With the
+        target sybil attacked the attacker can prevent the target from learning
+        that the other validators have \emph{not} failed. When those validators
+        are removed from the UNL due to the simulated failure the attacker does
+        not need to compromise/coerce as many validators as before. Equally, if
+        the target is using network split detection, the attacker can make the
+        non-compromised validators \emph{actually} fail as preparation for the
+        attack. Again, the lack of incentive to run a validator is problematic.
+
+\end{itemize}
 

### 2015-04-26 paper.tex
sp

@@
     \item The global consensus ledger has inherently poor $O(n^2)$ scaling;
-        even in the absense of a malicious attack the poor scalability of the
+        even in the absence of a malicious attack the poor scalability of the
         ledger is a threat to the viability of Ripple.

### 2015-04-26 paper.tex
Added attack flowchart and example

@@
 
-\section{Synergistic Attacks}
+\section{Attack flowchart}
 
-do a directed graph of likely synergistic attack sequences
+Figure \ref{fig:attack-flowchart} shows some of the stronger synergies between
+different attacks.
+
+Let's go through a full scenario, involving multiple attacks at once.  Ripple
+Labs has diversified their default UNL such that the $80\%$ of the validators
+are run by publicly known institutions evenly spread across the following five
+countries, United States, France, Switzerland, Russia, and China, with the
+remaining fifth being controlled by pseudoanonymous ``DarkWeb'' entities like
+The Silk Road Five, and Agora Market. As any one entity only comprises at most
+$1/6\text{th}$ of the UNL no single actor can block transactions.
+
+The Russian government has issued a court order seizing assets of Royal Dutch
+Shell, who has recently received funds from Naftogaz Ukraine to develop a new
+fracking project. Naftogaz in turn owes Russia for unpaid natural gas imports -
+quite a lot more than the funds Shell has, but simply punishing them for
+working with Naftogaz is a good start.
+
+By itself Russia is unable to freeze Shell's assets on the Ripple ledger
+through coercion of Russian-jurisdiction validators. So Russia begins with a
+transaction flood attack, slowly increasing load on the Ripple network to
+$10\text{MB}/\text{sec}$ of transactions. As a cover story a new
+gambling-oriented MMORPG is created that uses the Ripple ledger to record user
+accounts; Ripple Labs is happy to sell them the XRP to cover transaction fees.
+
+The transaction flood results in the pseudoanonymous ``DarkWeb'' validators
+quitting - the Tor network is too slow to handle the load, and they worry
+they're real identities will be discovered if they move their validator nodes
+to clearnet hosting.
+
+At this point Russia with $20\%$ UNL control could simply vote to freeze
+Shell's assets, achieving a denial-of-service. However, it's decided it would
+be better if the funds were actually recovered instead.
+
+Russia sets up a fake Chinese oil importer, hiring another Chinese company to
+do the actual transportation, and has them contact Shell asking to purchase
+some oil. They also find a wealthy Chinese investor with \$100 million USD
+worth of yuan in an account with a major Chinese bank, and has that investor
+``invest'' in the new importer. Meanwhile Russia's cyberwarfare team gets to
+work, succesfully compromising the Swiss and French validation nodes, and
+using the fact that they have agents in major Ukrainian ISPs to sybil attack
+the Ukrainian-based Shell office actually handling the sale. They can't
+compromise the US nodes - they're on a newer version - but they do happen to
+find a new zero-day consensus bug in the newer version.
+
+On the day of the sale unknown to Shell their Ukrainian office's internet
+connection is being filtered. The IT staff at their accounting office voices
+concern when their Ripple node monitoring tools say that the US validators just
+went off-line, although they're somewhat reassured when they hear news that the
+US nodes just hit a consensus issue. The captain says there's bad weather
+coming in, so they need to leave the port ASAP - with France and Switzerland
+approving the transaction they have confidence that the Free World has approved
+the money transfer.
+
+What's actually happened is Russian cyberwarfare experts created a fake
+consensus uses stolen Swiss and French private keys. Russia of course
+officially signed this fake consensus, as well as the real consensus, taking
+the position this was an above-the-board court ordered operation; China also
+signed, but doesn't officially say why. The wealthy Chinese investor
+conveniently claims to have suddenly found out about the ``fake'' oil importer
+and on the real Ripple ledger has the funds frozen by the Chinese bank backing
+them - not a big deal with the money still sitting in the original account he
+deposited it too, still under Chinese jurisdiction.
+
+Shell's Ukrainian office finally realises something is wrong after two hours of
+debugging, trying to figure out why an exchange wasn't seeing their transaction
+attempting to sell that yuan for USD. By that time the ship is in international
+waters, and is happy to keep steaming to China, where upon the oil is kept for
+the next ten years while the Chinese courts argue what to do with it.
+
+\begin{figure}
+    \centering
+    \includegraphics[scale=0.5]{figures/attack-flowchart.eps}
+    \caption{Attack flowchart}
+    \label{fig:attack-flowchart}
+\end{figure}
 

### 2015-05-05 paper.tex
Typo's
Thanks Tim Swanson!

@@
 not directly consume the output of other transactions. To prevent replay
-attacks accounts and transaction have sequence numbers; a transaction is only
+attacks, accounts and transactions have sequence numbers; a transaction is only
 valid if the Sequence number is exactly one greater than the last-valided
@@
 The actual process of determining what version of the ledger is correct starts
-with the Unique Node Listi, (UNL) a list of public keys meant to be associated
+with the Unique Node List, (UNL) a list of public keys meant to be associated
 with active nodes the node operator believes are
@@
 What's actually happened is Russian cyberwarfare experts created a fake
-consensus uses stolen Swiss and French private keys. Russia of course
+consensus using stolen Swiss and French private keys. Russia of course
 officially signed this fake consensus, as well as the real consensus, taking

### 2015-05-11 paper.tex
checkpoint
close to finished

@@
 \usepackage{mathtools}
-\usepackage{tikz}
 \usepackage{multirow}
-\usetikzlibrary{arrows,positioning}
 \begin{document}
-\title{Ripple Consensus Review}
+\title{Ripple Protocol Consensus Algorithm Review}
 \author{Peter Todd}
-\date{FIXME}
+\date{May 11th 2015}
 \maketitle
 
-\section{Activities performed}
+\section{Activities Performed}
 
@@
 
-    \item Reviewed whitepapers and development documentation at ripple.com
+    \item Reviewed whitepapers and development documentation at \url{https://ripple.com}.
 
-    \item Reviewed existing third-party critism.
+    \item Reviewed existing third-party criticism.
 
@@
 
-    \item Reviewed C++ implementation, specifically the 0.27.4 tag.\footnote{git commit 92812fe7239ffa3ba91649b2ece1e892b866ec2a}
+    \item Reviewed C++ implementation, specifically the 0.27.4 tag of
+        \texttt{rippled}.\footnote{git commit
+    \texttt{92812fe7239ffa3ba91649b2ece1e892b866ec2a} from \url{https://github.com/ripple/rippled}}
 
@@
 
-\section{Overall architecture}
+\section{Overall Architecture}
 
-Ripple has diverged significantly from the original concept of a decentralized
-network representing money explicitly as debt relationship between
-parties.\cite{btcmag-introducing-ripple} While the concept of a debt
-relationship still exists in the form of trust lines between participants, the
-bulk of the codebase and developer documentation now focuses\footnote{For
-    instance the ``Tutorials'' section of the Ripple website only explains how
-    to create transactions that modify the ledger; there is almost no information
-available on how to actually use the trustlines feature.} on the use and
-maintenance of a global ledger of transactions and account balances.
-Additionally a native currency, XRP, has been added, which is used for
-anti-spam transaction fees\footnote{A small amount of XRP is irrovocably
-destroyed for every transaction.}, as well as to serve as a universal currency.
+Ripple has diverged significantly from the original
+concept\cite{btcmag-introducing-ripple} of a decentralized network recording
+the transferral of money explicitly as debt relationships between parties.
+(Figure \ref{fig:orig-ripple}, showing entities A, B, and C participating in a
+transaction) While the concept of a debt relationship still exists in the form
+of trustlines between participants, the bulk of the codebase and developer
+documentation now focuses\footnote{For instance the ``Tutorials'' section of
+the Ripple website only explains how to create transactions that modify the
+ledger; there is almost no information available on how to actually use the
+trustlines feature.} on the use and maintenance of a global ledger of
+transactions and account balances.  Additionally a native currency, XRP, has
+been added, which is used for anti-spam transaction fees\footnote{A small
+amount of XRP is irrovocably destroyed for every transaction.} and to
+serve as a universal currency. (Figure \ref{fig:ripple-labs-ripple})
 
-\subsection{Hashing and serialization}
+\begin{figure}
+    \centering
+    \includegraphics[scale=0.3]{figures/orig-ripple.eps}
+    \caption{Original Ripple Architecture}
+    \label{fig:orig-ripple}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics[scale=0.3]{figures/ripple-labs-ripple.eps}
+    \caption{Ripple Labs Architecture}
+    \label{fig:ripple-labs-ripple}
+\end{figure}
+
+\section{Global Ledger}
+
+Figure \ref{fig:ledger-blockchain} shows the basic cryptographic structure of
+the Ripple ledger. As with Bitcoin, it consists of a set of chain of blocks,
+each consisting of a block header that commits to a previous block and a set of
+new transactions for that block. (via a merkle tree) Additionally each block
+header also commits to a merkelized trie\footnote{Oddly the trie is not binary,
+trading off performance for significantly larger proofs.} of the state of all
+account balances. The block chain doesn't itself deal directly with the
+consensus protocol. Rather validators sign messages as part of the consensus
+protocol, and knowledge of those messages lets Ripple nodes determine the
+correct chain.
+
+Developer documentation about exactly how the ledger is structured is somewhat
+spotty; the descriptions here are mostly based on direct analysis of the
+\texttt{rippled} source code.
+
+\begin{figure}
+    \centering
+    \includegraphics[scale=0.3]{figures/ledger-blockchain.eps}
+    \caption{Ripple Ledger Structure}
+    \label{fig:ledger-blockchain}
+\end{figure}
+
+\subsection{Transactions}
+
+Unlike Bitcoin, transactions increment and decrement account balances; they do
+not directly consume the output of other transactions. Transactions are also
+used to setup and modify trustlines. To prevent replay attacks, accounts and
+transactions have sequence numbers; a transaction is only valid if the Sequence
+number is exactly one greater than the last-valided transaction from the same
+account. Additionally an optional AccountTxnID field is available for use
+within transactions; the transaction is only valid if the sending account's
+previously-sent transaction matches the provided hash.
+
+Authentication is performed with a simple signature scheme; a scripting system
+is not available, nor is multisignature support. To implement the latter a
+highly complex single-purpose scheme\cite{ripple-wiki-multisign} has been
+proposed.
+
+\subsection{Low-Level Hashing and Serialization}
 
@@
 \begin{equation}
-    \textit{Serialize}(\text{obj}) = t_0 n_0 d_0 + \hdots + t_n n_n d_n
+    \textit{Serialize}(\text{obj}) = t_0+n_0+d_0 + \hdots + t_n+n_n+d_n
 \end{equation}
@@
 Hashing of objects generally uses the following scheme, implemented by
-STObject::getHash(), resulting in a $256\text{bit}$ digest:
+\texttt{STObject::getHash()}, resulting in a $256\text{bit}$ digest:
 
@@
 
-\subsection{Transactions}
+\subsection{Open Questions}
 
-Unlike Bitcoin, transactions increment and decrement account balances; they do
-not directly consume the output of other transactions. To prevent replay
-attacks, accounts and transactions have sequence numbers; a transaction is only
-valid if the Sequence number is exactly one greater than the last-valided
-transaction from the same account. Additionally an optional AccountTxnID field
-is available for use within transactions; the transaction is only valid if the
-sending account's previously-sent transaction matches the provided hash.
+\begin{itemize}
 
-Authentication is performed with a simple signature scheme; a scripting system
-is not available, nor is multisignature support. To implement the latter a
-highly complex single-purpose scheme\cite{ripple-wiki-multisign} has been
-proposed.
+    \item How exactly does the block header skiplist work? While the source code
+        refers to a skiplist, the actual block header format appears to only
+        allow block headers to refer to a single previous block header in a
+        linear fashion.
 
-\subsection{Ledger}
+    \item The \texttt{rippled} codebase suggests that in some circumstances account
+        balances can be negative. What are those circumstances exactly?
 
-Information on the exact structure of the Ripple ledger is somewhat spotty.
+    \item Do opportunities exist to ``shard'' the Ripple blockchain to improve
+        scaling by distributing load across parallel servers?
 
-how exactly does the skiplist work?
+    \item What plans exist (if any) to support SPV-style ``lite clients'' in
+        the Ripple protocol?
+
+\end{itemize}
+
+\section{Consensus Algorithm}
+
+Coming to consensus about the contents of the Ripple ledger is the role of the
+Ripple Protocol Consensus Algorithm\cite{ripple-consensus-paper}. In situations
+where the participants can be determined in advance, consensus algorithms are a
+well studied with a large variety of choices available. With that in mind while
+this review did some basic ``sanity checking'' of the specific algorithm, for
+the most part we assume it works correctly as described. Instead we'll focus on
+the assumptions surrounding the consensus algorithm, and how it can and will be
+used in practice.
 
@@
 
-The actual process of determining what version of the ledger is correct starts
-with the Unique Node List, (UNL) a list of public keys meant to be associated
-with active nodes the node operator believes are
+Like the vast majority of consensus algorithms, the Ripple consensus algorithm
+starts with a known set of nodes known to be participating in the consensus.
+This Unique Node List, (UNL) is a list of public keys meant to be associated
+with active (validating) nodes the node operator believes are
 ``unique''.\cite{ripple-wiki-unl} Ripple Labs suggests that UNL's ``should have
 100+ nodes on them.''\cite{ripple-wiki-unl} and provides a ``starter'' UNL at
-\url{https://ripple.com/ripple.txt} with $5$ to $8$ nodes.\footnote{Exactly how this UNL is
-generated is unknown; the author downloaded it on multiple occasions getting
-between $5$ to $8$ nodes each time, mostly the same.}
+\url{https://ripple.com/ripple.txt} with $5$ to $8$ nodes.\footnote{Exactly how
+this UNL is generated is unknown; the author downloaded it on multiple
+occasions getting between $5$ to $8$ nodes each time, mostly the same.}
 
-Through the Ripple Protocol Consensus Algorithm\cite{ripple-consensus-paper}
-nodes on the UNL vote to determine the contents of the Ripple ledger. While the
-actual protocol contains a number of rounds of proposals and voting the end
-result can be described as basically a supermajority vote: a transaction is
-only approved if $80\%$ of the UNL of a server agrees with
-it.\cite[3.2]{ripple-consensus-paper} Put another way, if $20\%$ of the UNL
-choose to reject a transaction, it will not be included in the ledger.
+Through the consensus algorithm nodes on the UNL vote to determine the contents
+of the Ripple ledger. While the actual protocol contains a number of rounds of
+proposals and voting the end result can be described as basically a
+supermajority vote: a transaction is only approved if $80\%$ of the UNL of a
+server agrees with it.\cite[3.2]{ripple-consensus-paper} Put another way, if
+$20\%$ of the UNL choose to reject a transaction, it will not be included in
+the ledger.
 
@@
 public validators, yet validation at minimum raises potential legal issues such
-as lawsuits for negligence. Again, the least-risk option is to not publicly
-validate the Ripple ledger.
+as lawsuits for negligence or aiding financial crimes. Again, the least-risk
+option is to not publicly validate the Ripple ledger.
 
@@
 transactions, or with more sophistication, the attacker can fork the Ripple
-ledger/blockchain and use the fork to doublespend. The attack is stopped by
+ledger/blockchain and use the fork to double-spend. The attack is stopped by
 identifying it, and then distributing software patches. Provided the Ripple
@@
 
-    \item The main implementation of the Ripple protocol, \emph{rippled}, does
+    \item The main implementation of the Ripple protocol, \texttt{rippled}, does
         not cleanly separate the consensus-critical part of the codebase from
@@
 
-    \item The rippled node implementation does not check full history. This
+    \item The \texttt{rippled} node implementation does not check full history. This
         means if validators sign an invalid ledger, the transactions that make
@@
 The attacker inserts a backdoor into a widely used Ripple protocol
-implementation. (e.g. the rippled node software) The attacker may be an
+implementation. (e.g. the \texttt{rippled} node software) The attacker may be an
 internal Ripple Labs employee, an external member of the public submitting a
 pull-request, or they may compromise the software actually downloaded by
-end-users by compromising binary or sourcecode hosting. (e.g. GitHub)
+end-users by compromising binary or source code hosting. (e.g. GitHub)
 
@@
     \item Ripple Labs appears to use GitHub - a third party -
-        as their master sourcecode repository, with development happening
+        as their master source code repository, with development happening
         through pull-requests on GitHub. While this does place trust in a
-        third-party, PGP signing sourcecode during development can mitigate
+        third-party, PGP signing source code during development can mitigate
         this risk. However there appears to be no PGP signatures on any git
@@
 
-    \item The rippled implementation doesn't yet appear to support separating
+    \item The \texttt{rippled} implementation doesn't yet appear to support separating
         the validation/consensus and signing functionality into different
@@
 
-What's actually happened is Russian cyberwarfare experts created a fake
+What has actually happened is Russian cyberwarfare experts created a fake
 consensus using stolen Swiss and French private keys. Russia of course
@@
 
+\section{Conclusions and future work}
+
+Nearly all the attack scenarios analysed above are direct outcomes of the need
+for consensus; the actual blockchain technology that records the ledger is
+relatively uninteresting. That the consensus requirement is for global
+consensus rather than the original Ripple's local consensus creates additional
+problems such as co-ordination of UNL's, incentives for such co-ordination, and
+unclear incentives for validating nodes to validate correctly or honestly.
+Finally the requirement of global consensus creates scalability concerns,
+privacy concerns, and jurisdictional concerns.
+
+A key question that should be answered in future work is if the goals of the
+Ripple system need global consensus at all? If global consensus can be avoided,
+or at least it's use minimized, many of these issues may go away.
+
 \bibliographystyle{plain}

### 2015-05-11 paper.tex
close to finished

@@
 
-    \item Reviewed whitepapers and development documentation at \url{https://ripple.com}.
+    \item Reviewed white papers and development documentation at \url{https://ripple.com}.
 
@@
 been added, which is used for anti-spam transaction fees\footnote{A small
-amount of XRP is irrovocably destroyed for every transaction.} and to
+amount of XRP is irrevocably destroyed for every transaction.} and to
 serve as a universal currency. (Figure \ref{fig:ripple-labs-ripple})
@@
 Authentication is performed with a simple signature scheme; a scripting system
-is not available, nor is multisignature support. To implement the latter a
+is not available, nor is multi-signature support. To implement the latter a
 highly complex single-purpose scheme\cite{ripple-wiki-multisign} has been
@@
 connectivity may have been associated with network latency. The paper does
-state that "nodes who’s latency grows larger than a preset bound $b$ are
-removed from all UNLs"\cite[3.4.1]{ripple-consensus-paper} but does not say if
+state that ``nodes whose latency grows larger than a preset bound $b$ are
+removed from all UNLs''\cite[3.4.1]{ripple-consensus-paper} but does not say if
 this removal is meant to be permanent\footnote{In the existing implementation
@@
         \multirow{3}{*}{Local} & $100\%$ & Consensus & DoS & DoS \\ \cline{2-5}
-        ~ & $50\%$ & DoS & DoS & Doublespend \\ \cline{2-5}
-        ~ & $10\%$ & Doublespend & Doublespend & Doublespend \\
+        ~ & $50\%$ & DoS & DoS & Double-spend \\ \cline{2-5}
+        ~ & $10\%$ & Double-spend & Double-spend & Double-spend \\
     \end{tabular}
@@
 
-\section{Attack scenarios}
+\section{Attack Scenarios}
 
@@
 
-The attacker creates large numbers of transactions - either covertely or
+The attacker creates large numbers of transactions - either covertly or
 overtly - until the Ripple network fails due to overload, or more likely,
@@
 short to medium term such an attack would quickly drive up that fee. A
-succesful attack requires deep pockets as the attacker must outspend other
+successful attack requires deep pockets as the attacker must out-spend other
 Ripple users.
@@
     \item A clever attacker wishing to overtly attack the Ripple network can
-        masquerade their traffic as legit economic activity; from the point of
+        masquerade their traffic as legit economic activity. From the point of
         view of Ripple Labs - a major XRP owner - such an attack may not even
@@
         attacks depending on the high centralization of the Ripple network as
-        even ``unsuccesful'' scalability attacks drive further centralization.
+        even ``unsuccessful'' scalability attacks drive further centralization.
         For instance, if the response to a scalability attack is for Ripple
@@
 
-\subsection{Coercion of validators}
+\subsection{Coercion of Validators}
 
@@
 
-Validators are coerced through poltical/legal/criminal means into attacking
+Validators are coerced through political/legal/criminal means into attacking
 some or all Ripple users by preventing valid transactions from entering the
@@
         attack would simply shut the Ripple network down for all users rather
-        than result in a succesful theft.
+        than result in a successful theft.
 
@@
         identification to use the Ripple network in the near future are an
-        example of a succesful coercion attack. It can be argued that Ripple
+        example of a successful coercion attack. It can be argued that Ripple
         Labs is increasing the risk of coercion attacks by simultaneously
@@
 
-\subsection{Software backdoor}
+\subsection{Software Backdoor}
 
@@
         HTTP; while source code is available through git, neither commits nor
-        release tags are signed in anyway. \textbf{This is a serious ommission
+        release tags are signed in anyway. \textbf{This is a serious omission
         that has lead to significant monetary losses in the past.} Ripple Labs
@@
 
-\subsection{Theft of validator secret keys}
+\subsection{Theft of Validator Secret Keys}
 
@@
 exploit, backdoor, compromised hosting company, etc. While not useful for the
-attacker in of itself, the stolen keys can be used for further atttacks such as
+attacker in of itself, the stolen keys can be used for further attacks such as
 DoS or a simulated ledger attack.
@@
     \item Due to the high centralization of UNL validators, a relatively small
-        number of validators need to be compromised for a succesful attack.
+        number of validators need to be compromised for a successful attack.
 
     \item Most validators are running nearly the same, if not the exact same,
-        software, making it likely for backdoor/exploit attack to succesfully
+        software, making it likely for backdoor/exploit attack to successfully
         exfiltrate the secret keys. Due to the high consensus-critical
@@
 
-\subsection{Simulated ledger}
+\subsection{Simulated Ledger}
 
@@
 sybil attacking the target the attacker can then perform double-spend attacks.
-Sybil attacks are relatively cheap, requiring just the aquisition of a large
+Sybil attacks are relatively cheap, requiring just the acquisition of a large
 number of IP addresses, e.g. via cloud hosting and/or botnet rental. In some
@@
 
-\section{Attack flowchart}
+\section{Attack Flowchart}
 
@@
 
-Let's go through a full scenario, involving multiple attacks at once.  Ripple
+Let's go through a full scenario involving multiple attacks at once.  Ripple
 Labs has diversified their default UNL such that the $80\%$ of the validators
@@
 
-At this point Russia with $20\%$ UNL control could simply vote to freeze
-Shell's assets, achieving a denial-of-service. However, it's decided it would
-be better if the funds were actually recovered instead.
+At this point Russia's $20\%$ UNL control could simply vote to freeze Shell's
+assets, achieving a denial-of-service. However, it's decided it would be better
+if the funds were actually recovered instead.
 
@@
 ``invest'' in the new importer. Meanwhile Russia's cyberwarfare team gets to
-work, succesfully compromising the Swiss and French validation nodes, and
+work, successfully compromising the Swiss and French validation nodes, and
 using the fact that they have agents in major Ukrainian ISPs to sybil attack
@@
 
-\section{Conclusions and future work}
+\section{Conclusions and Future Work}
 
@@
 Ripple system need global consensus at all? If global consensus can be avoided,
-or at least it's use minimized, many of these issues may go away.
+or at least its use minimized, many of these issues may go away.
 

### 2015-05-11 paper.tex
close to finished

@@
 
-The Ripple consensus review investigation had four major activities associated
-with it:
+Four major activities were performed during the review:
 
@@
             portions of Bitcoin Core codebase are currently being separated
-            into a stand-alone \emph{libconsensus} library.}
+            into a stand-alone \texttt{libconsensus} library.}
 

### 2015-05-11 paper.tex
close to finished
scale

@@
     \centering
-    \includegraphics[scale=0.3]{figures/orig-ripple.eps}
+    \includegraphics[scale=0.35]{figures/orig-ripple.eps}
     \caption{Original Ripple Architecture}
@@
     \centering
-    \includegraphics[scale=0.3]{figures/ripple-labs-ripple.eps}
+    \includegraphics[scale=0.35]{figures/ripple-labs-ripple.eps}
     \caption{Ripple Labs Architecture}
@@
 
-Developer documentation about exactly how the ledger is structured is somewhat
-spotty; the descriptions here are mostly based on direct analysis of the
+Detailed documentation about exactly how the ledger is structured is spotty;
+the descriptions here are mostly based on direct analysis of the
 \texttt{rippled} source code.
@@
     \centering
-    \includegraphics[scale=0.3]{figures/ledger-blockchain.eps}
+    \includegraphics[scale=0.35]{figures/ledger-blockchain.eps}
     \caption{Ripple Ledger Structure}

### 2015-05-11 paper.tex
close to finished

@@
 number is exactly one greater than the last-valided transaction from the same
-account. Additionally an optional AccountTxnID field is available for use
+account. Additionally an optional \texttt{AccountTxnID} field is available for use
 within transactions; the transaction is only valid if the sending account's

### 2015-05-12 paper.tex
fix typos

@@
 small business as it would on a $\$1,000,000,000/\text{day}$ multinational, the
-latter could lose orders of magnitude more money for what it at a technical
-level basically the same attack. A closely issue with attempting to predict
+latter could lose orders of magnitude more money for what is at a technical
+level basically the same attack. A closely related issue with attempting to predict
 monetary losses is that at some point losses can be sufficiently high that they
@@
 exploit, backdoor, compromised hosting company, etc. While not useful for the
-attacker in of itself, the stolen keys can be used for further attacks such as
+attacker in and of itself, the stolen keys can be used for further attacks such as
 DoS or a simulated ledger attack.
@@
 quitting - the Tor network is too slow to handle the load, and they worry
-they're real identities will be discovered if they move their validator nodes
+their real identities will be discovered if they move their validator nodes
 to clearnet hosting.
@@
 them - not a big deal with the money still sitting in the original account he
-deposited it too, still under Chinese jurisdiction.
+deposited it to, still under Chinese jurisdiction.
 
